Red Wine Analysis by Kalpana
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(reshape)
library(GGally)
library(RColorBrewer)
library(gridExtra)
library(dplyr)
library(knitr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine = read.csv('wineQualityReds.csv')
wine_duplicate = read.csv('wineQualityReds.csv')
```

# About the dataset:

   The following analysis is based on the dataset by
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. Modeling wine 
preferences by data mining from physicochemical properties. In Decision Support
Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
[Pre-press (pdf)] 
http://www3.dsi.uminho.pt/pcortez/winequality09.pdf [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib

You can read more on this from this text:
https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt


# Overview of the data

``` {r echo = FALSE,message = FALSE, DATASET_OVERVIEW1}
dim(wine)
```
There are about 1599 observations of 13 variables


``` {r echo = FALSE, DATASET_OVERVIEW2}
names(wine)
```

The type of variables are:

``` {r echo = FALSE, DATASET_OVERVIEW3} 
str(wine)
```

``` {r echo = FALSE, DATASET_OVERVIEW4} 
summary(wine)

```

# Data Wrangling:
 It seems that x is an index variable which we wouldn't need for analysis. So we
 can remove it. Also, the quality variable is numeric, we can convert it to 
 factor variable.
 
``` {r echo= TRUE,message = FALSE }
wine <-  wine[2:13]
wine$quality <- factor(wine$quality, ordered = TRUE)
levels(wine$quality)

```

# Univariate Plots Section

Let us plot the histograms of the individual variables to understand 
them better.

QUALITY:


``` {r echo = FALSE, message = FALSE,warning = FALSE,  Univariate_Plots}
ggplot(aes(x= quality), data  = wine) + 
  geom_histogram( fill = 'lightgreen', stat = 'count')  

summary(wine$quality)
```

Majority of the wines in the dataset have the quality rating of 5 and 6.

FIXED ACIDITY:

```{r echo=FALSE, message = FALSE, warning=FALSE, Univariate_Plots1}
ggplot(aes(x= fixed.acidity), data = wine) + geom_histogram(color = I('red'),
                                                        fill = I('burlywood'))

summary(wine$fixed.acidity)
```

The fixed acidity is the acidity of most acids involved with wine
which are non volatile. It has a normal distribution with few outliers.

CITRIC ACID:

```{r echo=FALSE, message = FALSE, warning=FALSE, Univariate_Plots2}
ggplot(aes(x= citric.acid ), data = wine) +
  geom_histogram(color = I('darkgreen'), fill = I('#56B4E9'))+
  scale_x_continuous(limits = c(0, 0.8), breaks = seq(0,0.8,0.2))  

table(wine$citric.acid)

```

citric acid is found in small quantities. citric acid can add 'freshness' and 
flavor to wines.There are many zero values for this variable. 

ALCOHOL:

The alcohol content is another important aspect to consider when we are 
purchasing wine:

```{r echo=FALSE, message = FALSE, warning=FALSE, Univariate_Plots3}
ggplot(aes(x= alcohol  ), data = wine) + 
  geom_histogram( color = I('green') , fill = I('grey') )  + 
  xlab(" % of alcohol content")

summary(wine$alcohol)
# determine % of wines with alcohol % between 9 and 11 inclusive
dim(wine[ which(wine$alcohol >= 9 & wine$alcohol <= 11), ] )[1] / 
  dim(wine)[1]


```

In almost 75% of the wines, percentage of alcohol content is around 9.5 to 11 .

RESIDUAL SUGAR AND CHLORIDES :

The residual sugar is the amount of sugar remaining after fermentation stops. 
It's rare to find wines with less than 1 gram/liter and wines with greater than
45 grams/liter are considered sweet. The wines in our dataset have maximum
15% residual sugar ,which means they are not sweet.


``` {r echo = TRUE, message = FALSE, warning = FALSE, Univariate_Plots4}
ggplot(aes(x=residual.sugar), data=wine) +
  geom_histogram(binwidth=0.05, fill='blue', colour='black') + 
  scale_x_continuous(breaks=0:16)
# This is has too many outliers. lets exclude the 95th percentile

ggplot(aes(x = chlorides), data = wine) + 
  geom_histogram(fill = 'yellow3', colour = 'black')


```


Plotting the histogram of residual sugar and chlorides, we see many outliers.
Excluded the 95th percentile and plotted again. I added a log scale to get a 
normalized distribution.

``` {r echo = TRUE, message = FALSE, warning = FALSE, Univariate_Plots_4a}
ggplot(wine, aes(x=residual.sugar)) + 
  geom_histogram(fill='blue', colour='black') +
  scale_x_log10(breaks=seq(1,6,1)) +
  xlim(0, quantile(wine$residual.sugar, c(0.95)))   
   
summary(wine$residual.sugar)

```



``` {r echo = TRUE, message = FALSE, warning = FALSE, Univariate_Plots5}
ggplot(aes(x = chlorides ), data =wine) +
  geom_histogram(fill='yellow3', colour='black') +
  scale_x_log10()+
  xlim(0.025, quantile(wine$chlorides, c(0.95))) 
  
```

# Univariate Analysis


### What is the structure of your dataset?

There are 1599 observations of 13 variables

### What is/are the main feature(s) of interest in your dataset?
Quality is the main feature of interest.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Fixed acidity, citric acid, residual sugar,alcohol,chlorides are also features
to be considered while investigating the quality.

### Did you create any new variables from existing variables in the dataset?
Yes I will be creating one new variable for categorising the quality rating.

### Of the features you investigated, were there any unusual distributions? \

Citric acid had unusual distribution as there are many with 0 as the value
(about 132 instances).

Residual sugar and chlorides had the maximum outliers. 

###Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Not many changes need to be done as the dataset is already tidy. I removed the 
index variable X as it is not necessary for the analysis.I also changed the 
quality variable to factor so that we can plot each of them easily.

# Bivariate Plots Section



```{r echo=TRUE, message=FALSE, warning=FALSE, Bivariate_Plots1}

ggcorr(wine,label= TRUE, hjust = 0.9,layout.exp = 1, label_round = 2)


# this shows the corr coeff of variables. so we can decide which ones to 
#concentrate on. 

```

Points to consider:

Quality has strong correlation with the following:

*alcohol:quality = 0.48

*sulphates:quality = 0.25

*citric.acid:quality = 0.23

*fixed.acidity:quality = 0.12


Let us plot them for better understanding.

``` {r echo=FALSE,message = FALSE, warning = FALSE,  Bivariate_Plots2}


library(grid)

q1 <- ggplot( aes(x=alcohol ,  color = quality ),data = wine) + 
  geom_density(alpha = 0.5)+ xlab ('alcohol (% by volume)')
  
q2 <- ggplot( aes(y = citric.acid, x =quality), data = wine) + 
  geom_boxplot() +xlab('citric acid in g / dm^3')
  
q3 <- ggplot( aes(y= fixed.acidity, x =quality), data = wine) + 
  geom_boxplot()+ ylab('fixed acidity(g / dm^3)')
q4 <- ggplot(wine, aes(x=sulphates, color= quality)) + 
  geom_density(alpha=0.2) + xlim(0.25, quantile(wine$sulphates,c(0.99)))+
  xlab ('sulphates in g / dm^3')

grid.arrange(q1,q2,q3,q4, ncol = 2,
top = textGrob('Relationship of alcohol, sulphates, fixed and volatile acidity
               on Wine quality', gp=gpar(fontsize=20,font=3))  ) 


```

Fixed acidity and citric acid have strong positive correlation with quality.

As we saw in univariate analysis, it is evident that quality increases as
the percentage of alcohol increases. The summary statistics for alcohol:

``` {r echo=FALSE,message = FALSE, warning = FALSE,  Bivariate_Plots3}
by(wine$alcohol, wine$quality, summary)
```



It is surprising to see sulphates,a wine additive which can contribute to
sulfur dioxide gas (S02) levels has so much to do with quality. I later learnt
that sulphates do act as an antimicrobial and antioxidant and thats what 
makes it so important.As concentration of sulphate increases, quality does 
increase. The summary statistics :


``` {r echo= FALSE,message = FALSE, warning = FALSE,  Bivariate_plots4}
by (wine$sulphates, wine$quality, summary)
```

These variables have negative correlation with quality:

*volatile.acidity:quality = -0.39

*total.sulfur.dioxide:quality = -0.19

*density:quality = -0.17

*chlorides:quality = -0.13
  
Note the highest negative correlation is with volatile acidity. All acids in 
wine are non- volatile except for acetic acid which is volatile.
volatile acidity is the amount of acetic acid in wine, which at too high of 
levels can lead to an unpleasant, vinegar taste. 

``` {r echo=FALSE,message = FALSE, warning = FALSE,  Bivariate_Plots5}
ggplot(aes( x= volatile.acidity, color = quality) , data = wine) + 
  geom_density()
  
```

Looking at the correlation between other variables, we note :

*fixed.acidity:citric.acid = 0.67

*fixed.acidity:density = 0.67

*fixed.acidity:pH = -0.68

*free.sulfur.dioxide:total.sulfur.dioxide = 0.67

*density:alcohol = -0.50

*citric.acid:pH = -0.54
  
  
pH describes how acidic or basic a wine is on a scale from 0 (very acidic) to
14 (very basic). Thus when citric acid increases, pH decreases indicating that
the wine is getting more acidic. Fixed acidity includes acidity of citric acid 
also, as it is a non volatile acid. Obviously citric acid and fixed acidity are 
positively correlated. Thus pH decreases with increase in fixed acidity. 

On Plotting the relationship between fixed acidity and other variables, we can 
understand the pattern in detail.


``` {r echo=FALSE ,message = FALSE, warning = FALSE,  Bivariate_Plots6}
p1 <- ggplot(aes(x= fixed.acidity, y= density) , data = wine) + 
  geom_point(alpha=0.2,color = I('green')) +  
  stat_smooth(method = "lm")
p2 <-ggplot(aes(x= fixed.acidity, y= citric.acid) , data = wine) + 
  geom_point(alpha=0.2,color = I('green')) +  
  stat_smooth(method = "lm")
p3 <-ggplot( aes(x=fixed.acidity, y= pH), data = wine) + 
  geom_jitter(alpha=0.2,color = I('coral3')) + 
  stat_smooth(method = "lm")  
p4 <- ggplot(aes(x= citric.acid, y= pH) , data = wine) + 
  geom_point(alpha=0.2,color = I('coral3')) +  
  stat_smooth(method = "lm")

grid.arrange(p1,p2,p3,p4,ncol=2)
```


Fixed acidity is positively correlated with density and citric acid . Both
fixed acidity and citric acid have inverse relationship with pH.


Similarly, total sulfur dioxide is the amount of free and bound forms of S02.
Hence, when free sulfur dioxide increases, total sulfur dioxide also 
increases.


``` {r echo=FALSE,message = FALSE, warning = FALSE,  Bivariate_Plots7}
ggplot( aes(free.sulfur.dioxide,total.sulfur.dioxide), data = wine) + 
  geom_jitter(alpha=0.2) + 
  stat_smooth(method = "lm")
```

One other important relationship to note is the negative correlation of density 
and alcohol.


``` {r echo=FALSE,message = FALSE, warning = FALSE,  Bivariate_Plots8}
ggplot( aes(alcohol,density), data = wine) + 
  geom_jitter(alpha=0.2,color = I('brown')) + 
  stat_smooth(method = "lm")
```



# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Quality is positively correlated to alcohol, sulphates, fixed acidity.
It is neagtively correlated to volatile acidiity.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Yes. Fixed acidity had positive correlations with density and citric acid.
whereas, it exhibited inverse relationship with pH.

### What was the strongest relationship you found?

Alcohol exhibited strongest relationship with quality. When comparing other 
variables, fixed acidity and pH had the strongest negative correlation.

# Multivariate Plots Section




The main factors which determine the quality are alcohol and sulphates. Lets
plot them to see their relationships.


```{r echo=FALSE,message = FALSE, warning = FALSE,  Multivariate_Plots1}
ggplot( aes(x = sulphates, y = alcohol, colour=quality), data = wine) +
  scale_x_continuous(limits =c(0.25,1.5), breaks = seq(0.25,1.5,0.25) ) +
 geom_density_2d(bins= 2) +
 geom_point(alpha = 0.1, position = 'jitter')


```


In the above plot, higher quality wines are at the top right and lower quality 
wines are at the bottom left. 
Though we can plot various variables against quality, Let us categorise the
quality ratings so as to get a more clear picture. The quality ratings are from
3 to 8.

Let us categorise them as low, average, good and high quality wines based on the 
rating.

```{r echo= TRUE, warning= FALSE, Multivariate_Plots2}
wine <- mutate(wine, rating = 
                 ifelse(quality < 5, 'Low', 
                   ifelse(quality == 5 , 'Average', 
                   ifelse(quality == 6 , 'Good', 'High'))))
wine$rating <- factor(wine$rating, 
                             levels=c("Low", "Average", "Good", "High"))

summary(wine$rating)
```

Now let us replot the earlier plot with rating variable.


```{r echo=FALSE, warning= FALSE, Multivariate_Plots3}
ggplot( aes(x = sulphates, y = alcohol,color = quality), data = wine)+
  facet_wrap(~ rating)+
  geom_point(alpha =0.2) +
 geom_smooth(method=lm, se = FALSE) 
  
    

```


High quality wines have more alcohol and less sulphates. In average and good 
quality wines, the percentage of alcohol and sulphates is kind of balanced.
In low quality wines, both are equally high.

```{r echo=FALSE, warning= FALSE, Multivariate_Plots4}
ggplot( aes(x = log10(sulphates), y = citric.acid, color = rating), data = wine)+
  facet_wrap(~ rating)+
  geom_point() +
  geom_smooth(method = 'lm')



```

Plotting density and alcohol


```{r echo=FALSE, warning= FALSE, Multivariate_Plots5}
ggplot( aes(x = density, y = alcohol,color = quality), data = wine)+
  facet_wrap(~ rating)+
  geom_point() +
 geom_smooth(method=lm, se = FALSE) 
    
```

With increase in percentage of alcohol, density decreases. High quality wines
have higher percentage of alcohol in them, but may be with a fine balance with 
the density.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

I observed that  lower density and higher alcohol content gives a good wine.
Higher quantity of sulphates and citric.acid makes a good wine too
but with lower volatile acidity only.

### Were there any interesting or surprising interactions between features?

Volatile acidity, fixed acidity and pH must be finely balanced to get the best 
quality wine.


### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
NO.

------

# Final Plots and Summary



### Plot One

```{r echo=FALSE, message = FALSE , warning= FALSE, Plot_One}
ggplot(aes(x= citric.acid ), data = wine) +
  geom_histogram(color = I('darkgreen'), fill = I('#56B4E9'))+
  scale_x_continuous(limits = c(0, 0.8), breaks = seq(0,0.8,0.2))  +
  labs(x= 'citric acid (g / dm^3)')+
  ggtitle('Distribution of Citric acid in wines')

```


### Description One

Citric acid has a nice positive correlation with quality(0.23). So I plotted it 
understand it better. To my surprise, very very small quantity (< 1 g) of citric
acid was added to wine. Many wines didnot have citric acid at all (0 g/ltr).
I thought it was because of inadequate data. Deeper analysis revealed there are
many wines which doesnt have citric acid in it (according to tests conducted by
the Food Standards Australia New Zealand website). The data and the analysis 
of our dataset reveals the same result.


### Plot Two
```{r echo=FALSE,message = FALSE, warning = FALSE, Plot_Two}

library(grid)

plot1 <- ggplot( aes(x=alcohol ,  color = quality ),data = wine) + 
  geom_density(alpha = 0.5)+ xlab ('alcohol (% by volume)')
  
plot2 <- ggplot( aes(y = volatile.acidity, x =quality), data = wine) + 
  geom_boxplot() + ylab('volatile acidity(g / dm^3)')
  
plot3 <- ggplot( aes(y= fixed.acidity, x =quality), data = wine) + 
  geom_boxplot()+ ylab('fixed acidity(g / dm^3)')
plot4 <- ggplot(wine, aes(x=sulphates, color= quality)) + 
  geom_density(alpha=0.2) + xlim(0.25, quantile(wine$sulphates,c(0.99)))+
  xlab ('sulphates in g / dm^3')

grid.arrange(plot1,plot2,plot3,plot4, ncol = 2,
top = textGrob('Relationship of alcohol, sulphates,fixed and volatile
acidity on Wine quality', gp=gpar(fontsize= 15,font=3))  ) 




```

### Description Two
The main factors which determine the quality of the wine have been plotted.
Percentage of alcohol, fixed acidity and sulphates have positive effect on 
quality. Wine quality and volatile acidity are negatively correlated.
In all cases, we see outliers in quality ratings 5 and 6 , showing there may be 
other factors that influence quality.


### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot( aes(x = volatile.acidity, y = alcohol, color = quality), data = wine)+
  facet_wrap(~rating)+
  geom_point(alpha =0.2) +
  geom_smooth(method=lm, se = FALSE) +
  ggtitle('Relationship between volatile acidity, alcohol and quality')+
  labs(x= 'volatile acidity in g / dm^3', y = 'alcohol (% by volume)')
  
    

```

### Description Three

The two strong positively correlated(alcohol) and negatively correlated(volatile.
acidity) factors have been plotted against quality. It is understood that high 
alcohol and low volatile acidity might give higher quality wines.
Please note that volatile acid , acetic acid can contribute to wine fault , 
which might make the wine undrinkable.


------

# Reflection

  This red wine dataset has 1599 observations and 13 variables.This is a 
perfectly tidy and small dataset to be used for EDA by beginners.The step-by-
step approach from univariate to bivariate and then to multivariate plots
was a nice way to explore a dataset. It is clear from the outset , 
the main factors that determine the quality. but plotting the correlations
revealed more insights.I guessed alcohol would be one of the main contributor.
Surprisingly, sulphates and acetic acid do have roles to play. Citric acid,
though correlated well, was added in very less quantities and doesnot have much
effect except for the freshness in wines.

  Future work would be to build a model to predict the wine quality. This 
dataset had more wines with rating 5 and 6. May be the next step would be to 
get more data on wines with other quality ratings too.

  The main struggle I faced was finding an apt dataset and tidying it. I was 
initially browsing thro various data and chose a few. Looking in detail, 
there were many NAs, repetitions and a whole big bunch of data, which I did not 
know how to interpret. As it was taking way too much of time, I resented to
one of the suggested datasets. It was worth a decision. I was able to learn a 
lot from this red wine dataset. Had I taken some other big dataset, I would 
have focussed on tidying rather than exploring the dataset.


